### 2005-10-17 Documentation/part1.tex
- Added documentation for user-defined systematics functions (these changes   affect the chapters "Calculating chi^2 with systematics only" and   "Experiment definitions with AEDL -> Rules and the treatment of systematics" - Changed the names of several functions in user_chi.c to give them more   unambiguous names (glbGetSignalBin -> glbGetSignalBinInRule, glbGetBackgroundBin   -> glbGetBackgroundBinInRule, glbGetMeasuredBin -> glbGetObservedBinInRule). JK

@@

Keeping all oscillation parameters and matter density scaling factors fixed,
 one can use the following functions to obtain the total $\chi^2$ of all 
 specified oscillation channels including systematics:
\begin{function} 
\GLBNS{ChiSys}
{\tt double glbChiSys(const glb\_params in,int exp, int rule)} returns
the $\chi^2$ for the (fixed) oscillation parameters {\tt in}, the
experiment number {\tt exp}, and the rule number {\tt rule}. For all
experiments or rules, use \GLBC{GLB\_ALL} as parameter value.
\end{function}
Note that the result of {\tt glbChiSys} for all experiments or rules
corresponds to the sum of all of the individual {\tt glbChiSys} calls. 
This equality will not hold for the minimizers in the next sections anymore. 
 An example how to use  {\tt glbChiSys} can be found on page~\pageref{ex:corrth13dcp}.

@@

>>>>

\section{Built-in systematics}
\index{norm}{Systematics!builtin functions}

@@

Keeping all oscillation parameters and matter density scaling factors fixed,
 one can use the following builtin functions to obtain the total $\chi^2$ of all 
 specified oscillation channels including systematics:
\begin{function} 
\GLBNS{ChiSys}
{\tt double glbChiSys(const glb\_params in,int exp, int rule)} returns
the $\chi^2$ for the (fixed) oscillation parameters {\tt in}, the
experiment number {\tt exp}, and the rule number {\tt rule}. For all
experiments or rules, use \GLBC{GLB\_ALL} as parameter value.
\end{function}
Note that the result of {\tt glbChiSys} for all experiments or rules
corresponds to the sum of all of the individual {\tt glbChiSys} calls. 
This equality will not hold for the minimizers in the next sections anymore. 
 An example how to use  {\tt glbChiSys} can be found on page~\pageref{ex:corrth13dcp}.

@@

\section{User-defined systematics calculation}
\index{norm}{Systematics!user-defined}
\label{sec:userchi}

For some experimental setups the built-in systematics functions are not sufficient
because they cannot handle systematical errors that are correlated between different
parts of a multi-detector or multi-source setup. Therefore \GLOBES\ allows the user
to override the default $\chi^2$ calculation. The reader who wants to use this
feature should be familiar with \Sec~\ref{sec:channel} and~\Sec~\ref{sec:rules}
of the \AEDL\ manual.

To specify a user-defined $\chi^2$ algorithm, the following function can be used:
\begin{function}
\GLBNS{SetUserChi}
{\tt void glbSetUserChi(int exp,int rule,double (*chi\_func)(double x[]),int dim,%
double centers[],double errors[],char *info)} tells \GLOBES\ to use {\tt chi\_func}
to calculate $\chi^2$ for experiment {\tt exp} and rule {\tt rule} (these parameters
can take the value \GLBC{GLB\_ALL} to specify all experiments or all rules). {\tt dim} is
the number of variables parametrizing the systematical errors, while {\tt centers[]}
and {\tt errors[]} give their central values and standard deviations, respectively.
When \GLOBES\ calls {\tt chi\_func}, the parameter array {\tt x[]} will contain
the values of the systematics parameters for which $\chi^2$ should be calculated.
Deviating from the standard conventions, the entries of {\tt x[]} will start at
index 1.
\end{function}

In addition to calling {\tt glbSetUserChi}, {\tt @errordim\_sys\_on} has to be set
to 20 in the \AEDL\ file (see~\Sec~\ref{sec:rules}).

Note that {\tt chi\_func} will be called many times by the \GLOBES\ minimizers.
Therefore the function should be as efficient as possible. The function
will need low-level access to internal \GLOBES\ data. Therefore the following
routines are provided:
\begin{function}
\GLBNS{GetCurrentExp}
{\tt int glbGetCurrentExp()} gives the number of the experiment for which
{\tt chi\_func} has been called.
\end{function}
\begin{function}
\GLBNS{GetCurrentRule}
{\tt int glbGetCurrentRule()} gives the number of the rule in the current experiment
for which {\tt chi\_func} has been called.
\end{function}
\begin{function}
\GLBNS{GetObservedBinInRule}
{\tt double glbGetObservedBinInRule(int exp,int rule,int bin)} returns the number of
signal events for experiment {\tt exp}, rule {\tt rule} and bin {\tt bin}, as calculated
in the preceding call to {\tt glbSetRates}.
\end{function}
\begin{function}
\GLBNS{GetSignalBinInRule}
{\tt double glbGetSignalBinInRule(int exp,int rule,int bin)} returns the number of
signal events for experiment {\tt exp}, rule {\tt rule} and bin {\tt bin}, calculated
for the oscillation parameter values that are currently being tested.
\end{function}
\begin{function}
\GLBNS{GetBackgroundBinInRule}
{\tt double glbGetBackgroundBinInRule(int exp,int rule,int bin)} returns the number of
background events for experiment {\tt exp}, rule {\tt rule} and bin {\tt bin}, calculated
for the oscillation parameter values that are currently being tested.
\end{function}
\begin{function}
\GLBNS{GetChannelBin}
{\tt double glbGetChannelBin(int exp,int channel,int bin)} returns the number of
signal events for experiment {\tt exp}, channel {\tt channel} and bin {\tt bin}, calculated
for the oscillation parameter values that are currently being tested.
\end{function}
\begin{function}
\GLBNS{ShiftSignalEnergyScale}
{\tt int glbShiftSignalEnergyScale(int exp,int rule,double amount)} applies an energy
calibration error of magnitude {\tt amount} to the signal event rates in experiment
{\tt exp} and rule {\tt rule}. Subsequent calls to {\tt glbGetSignalBinInRule} will
return the modified event rates. For details on energy calibration errors, see
\Sec~\ref{sec:rules}.
\end{function}
\begin{function}
\GLBNS{ShiftBackgroundEnergyScale}
{\tt int glbShiftBackgroundEnergyScale(int exp,int rule,double amount)} applies an energy
calibration error of magnitude {\tt amount} to the background event rates in experiment
{\tt exp} and rule {\tt rule}. Subsequent calls to {\tt glbGetBackgroundBinInRule} will
return the modified event rates. For details on energy calibration errors, see
\Sec~\ref{sec:rules}.
\end{function}

Typical applications for user-defined systematics are reactor neutrino experiments,
where systematics play a crucial role. For a setup with near and far detectors,
the formula for $\chi^2$ is
\begin{align}
  \chi^2 =& \sum_{i=1}^{\textrm{\# of bins}} \sum_{d = N,F}2
     \left( O_{d,i} - (1 + a_R + a_d) T_{d,i} +
       (1 + a_R + a_d) T_{d,i} \log \frac{(1 + a_R + a_d) T_{d,i}}{O_{d,i}}  \right) \nonumber\\
         & \ \ + \frac{a_R^2}{\sigma_R^2} + \frac{a_N^2}{\sigma_N^2} + \frac{a_F^2}{\sigma_F^2}
\label{equ:chi2reactor}
\end{align}
where $O_{N,i}$ and $O_{F,i}$ are the event rates for the $i$-th bin in the near and
far detector, calculated for the assumed ``true'' values of the oscillation parameters.
$T_{d,i}$ are the event rates for the parameter values that are currently being tested.
$a_R$, $a_N$ and $a_F$ parametrize the small uncertainties in the reactor flux and the fiducial
mass of the two detectors. In this example their central values are assumed to be zero,
while their standard deviations are $\sigma_R$, $\sigma_N$ and $\sigma_F$.

The first line of \equ{chi2reactor} is the standard $\chi^2$ expression for
the Poisson distribution, while the terms in the second line are penalties
for deviations of the systematics parameters from their central values.

\example{Systematics for a reactor setup with near and far detector}{
The following code fragment calculates $\chi^2$ for a reactor setup with
a far detector (experiment 0) and a near detector (experiment 1),
including six different systematical errors: the flux normalization
of the reactor ({\tt x[1]}), the fiducial mass uncertainties of the far
({\tt x[2]}) and near ({\tt x[3]}) detector, the normalization of the
background ({\tt x[4]}) and energy calibration errors for the far
({\tt x[5]}) and near ({\tt x[6]}) detectors. We assume that the central
values and errors of these parameters are stored in the global arrays
{\tt sys\_centers} and {\tt sys\_errors}. The calculation follows
\equ{chi2reactor}.

\begin{quote}{\tt \footnotesize
  double chi2nearfar(double *x)                                                 \\
  \{                                                                            \\
\hspace*{0.5 cm}      double chi2 = 0;                                          \\
\hspace*{0.5 cm}      double chirate, measured\_rate;                           \\
\hspace*{0.5 cm}                                                                \\
\hspace*{0.5 cm}      glbShiftSignalEnergyScale(0, 0, x[5]);   /* Apply E calib. error */ \\
\hspace*{0.5 cm}      glbShiftSignalEnergyScale(1, 0, x[6]);                    \\
\hspace*{0.5 cm}      glbShiftBackgroundEnergyScale(0, 0, x[5]);                \\
\hspace*{0.5 cm}      glbShiftBackgroundEnergyScale(1, 0, x[6]);                \\
\hspace*{0.5 cm}      for (int i=0; i < NBINS; i++) \{                          \\
\hspace*{1.0 cm}        /* Statistical part of chi\textasciicircum2 for far detector */ \\
\hspace*{1.0 cm}        chirate = (x[1]+x[2]) * glbGetSignalBinInRule(0,0,i) +  \\
\hspace*{2.0 cm}                        (x[4]+x[3]) * glbGetBackgroundBinInRule(0,0,i);\\
\hspace*{1.0 cm}        measured\_rate = glbGetObservedBinInRule(0,0,i);        \\
\hspace*{1.0 cm}        chi2 += 2 * (measured\_rate - chirate);                 \\
\hspace*{1.0 cm}        if (measured\_rate > 0  \&\&  chirate > 0)              \\
\hspace*{1.5 cm}          chi2 += 2 * chirate * log(chirate / measured\_rate);  \\
\hspace*{1.0 cm}                                                                \\
\hspace*{1.0 cm}        /* Statistical part of chi\textasciicircum2 for near detector */ \\
\hspace*{1.0 cm}        chirate = (x[1]+x[3]) * glbGetSignalBinInRule(1,0,i) +  \\
\hspace*{2.0 cm}                        (x[4]+x[3]) * glbGetBackgroundBinInRule(1,0,i);\\
\hspace*{1.0 cm}        measured\_rate = glbGetObservedBinInRule(1,0,i);        \\
\hspace*{1.0 cm}        chi2 += 2 * (measured\_rate - chirate);                 \\
\hspace*{1.0 cm}        if (measured\_rate > 0  \&\&  chirate > 0)              \\
\hspace*{1.5 cm}          chi2 += 2 * chirate * log(chirate / measured\_rate);  \\
\hspace*{0.5 cm}      \}                                                        \\
\hspace*{0.5 cm}                                                                \\
\hspace*{0.5 cm}      for (int j=0; j < 6; j++)  /* Penalties */                \\
\hspace*{1.0 cm}        chi2 += square((x[j+1] - sys\_centers[j]) / sys\_errors[j]);\\
\hspace*{0.5 cm}                                                                \\
\hspace*{0.5 cm}      /* The function is called for each exp ->
                                       Assign chi\textasciicircum2/2 to each */ \\
\hspace*{0.5 cm}      return chi2 / 2;                                          \\
  \}                                                                            
}
\end{quote}
}

### 2005-10-17 Documentation/part2.tex
- Added documentation for user-defined systematics functions (these changes   affect the chapters "Calculating chi^2 with systematics only" and   "Experiment definitions with AEDL -> Rules and the treatment of systematics" - Changed the names of several functions in user_chi.c to give them more   unambiguous names (glbGetSignalBin -> glbGetSignalBinInRule, glbGetBackgroundBin   -> glbGetBackgroundBinInRule, glbGetMeasuredBin -> glbGetObservedBinInRule). JK

The user has the possibility to choose the set $\{\zeta_i\}$ of nuisance 
parameters which are minimized over. This choice is specified with the 
error dimension variable\index{norm}{Error dimension}, and the different 
possibilities are shown in
\Tab~\ref{tab:error_dim}. Since the error dimension defines the treatement
of systematics it is useful to have define a matched pair of error dimensions
for each rule, where on value describes how the event rate is computed
for \emph{no} systematics and the other one is with systematics 
(see also~\Sec~\ref{sec:systematics}).
The error dimensions for the case of no systematics is set for each rule 
by the value of {\tt @errordim\_sys\_off}\index{aedl}{{\tt rule}!errordim 
sys off@{\tt "@errordim\_sys\_off}}, whereas the error dimension
for systematics on is given by {\tt @errordim\_sys\_on}
\index{aedl}{{\tt rule}!errordim sys on@{\tt "@errordim\_sys\_on}}. Thus,
 for example,
the complete error dimension definition could look like  
\begin{quote}
{\tt 
\tb @errordim\_sys\_off = 2\\
\tb @errordim\_sys\_on = 0
}
\end{quote}
It is foreseen to add the possiblity to extend the set 
of error dimensions or the set of possible systematical errors in one
of the next versions of \GLOBES.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{center}
\begin{table}[t!]
\begin{center}
\begin{tabular}[h]{|c|cccc|c|l|}
\hline
Error dimension&$a$&$b$&$c$&$d$&Tilt/Calibration&Remarks\\
\hline
\hline
0&+&+&+&+&T&Systematics with tilt\\
2&-&-&-&-&-&No systematics\\
4&+&+&+&+&T&Total rates\\
7& $\infty$ &-& $\infty$ &-&-&Spectrum only\\
8&-&-&-&-&-&Total rates, No systematics\\
9&+&+&+&+&C&Systematics with calibration\\
\hline
\end{tabular}
\caption[Table of error dimensions]{\label{tab:error_dim}
Possible values of the error dimensions variable in \GLOBES\ and their meaning. If a parameter is designated with $+$, it will
be marginalized over, and therefore the corresponding error needs to 
have a non-zero value. If the cases with ``total rates'' in the remarks, 
the summation over the bins is performed \emph{before} computing 
the $\chi^2$, \ie, no spectral information is used. 
The error dimension~7 ``spectrum only'' leaves the 
normalization free ($\sigma_a=\sigma_c=\infty$), and therefore 
only the spectral information is used. As a consequence,
the settings for the normalization error will be ignored (designated 
with the symbol $\infty$).

>>>>

The user has the possibility to choose the set $\{\zeta_i\}$ of nuisance 
parameters which are minimized over. This choice is specified with the 
error dimension variable\index{norm}{Error dimension}, and the different 
possibilities are shown in
\Tab~\ref{tab:error_dim}. Since the error dimension defines the treatement
of systematics it is useful to have define a matched pair of error dimensions
for each rule, where on value describes how the event rate is computed
for \emph{no} systematics and the other one is with systematics 
(see also~\Sec~\ref{sec:systematics}).
The error dimensions for the case of no systematics is set for each rule 
by the value of {\tt @errordim\_sys\_off}\index{aedl}{{\tt rule}!errordim 
sys off@{\tt "@errordim\_sys\_off}}, whereas the error dimension
for systematics on is given by {\tt @errordim\_sys\_on}
\index{aedl}{{\tt rule}!errordim sys on@{\tt "@errordim\_sys\_on}}. Thus,
 for example,
the complete error dimension definition could look like  
\begin{quote}
{\tt 
\tb @errordim\_sys\_off = 2\\
\tb @errordim\_sys\_on = 0
}
\end{quote}
If none of the built-in algorithms listed in \Tab~\ref{tab:error_dim} is
suitable for a certain setup, the user can override them by setting
{\tt @errordim\_sys\_on} to 20 and using the \GLB{SetUserChi}
function to specify a specialized function for the calculation
of $\chi^2$. This function will then be called by the \GLOBES\ minimizers.
For details, see \Sec~\ref{sec:userchi}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{center}
\begin{table}[t!]
\begin{center}
\begin{tabular}[h]{|c|cccc|c|l|}
\hline
Error dimension&$a$&$b$&$c$&$d$&Tilt/Calibration&Remarks\\
\hline
\hline
0&+&+&+&+&T&Systematics with tilt\\
2&-&-&-&-&-&No systematics\\
4&+&+&+&+&T&Total rates\\
7& $\infty$ &-& $\infty$ &-&-&Spectrum only\\
8&-&-&-&-&-&Total rates, no systematics\\
9&+&+&+&+&C&Systematics with calibration\\
20&?&?&?&?&?&User-defined systematics\\
\hline
\end{tabular}
\caption[Table of error dimensions]{\label{tab:error_dim}
Possible values of the error dimensions variable in \GLOBES\ and their meaning. If a parameter is designated with $+$, it will
be marginalized over, and therefore the corresponding error needs to 
have a non-zero value. If the cases with ``total rates'' in the remarks, 
the summation over the bins is performed \emph{before} computing 
the $\chi^2$, \ie, no spectral information is used. 
The error dimension~7 ``spectrum only'' leaves the 
normalization free ($\sigma_a=\sigma_c=\infty$), and therefore 
only the spectral information is used. As a consequence,
the settings for the normalization error will be ignored (designated 
with the symbol $\infty$).
